---
title: "Lizard Wildflowers"
author: "Ben Holt"
date: "2026-04-14"
output: html_document
---
# 🌼 Wildflower Recording and Distribution on the Lizard Peninsula  
*An analysis of iNaturalist data using 1 km grid mapping*

---

## 📍 Introduction

This project began with a simple question: **where are the best places to go to look for wildflowers on the Lizard Peninsula?**

The [Lizard Peninsula](https://en.wikipedia.org/wiki/Lizard_Peninsula "Lizard Peninsula") is one of the most botanically rich parts of the UK, with a distinctive mix of coastal grassland, heathland and maritime habitats. Its mild climate and unusual geology support a wide range of flowering plants, making it an ideal place to explore patterns of plant diversity and distribution.

Rather than relying only on guidebooks or local knowledge, this report uses records from [iNaturalist](https://www.inaturalist.org "iNaturalist") to explore how flowering plants have actually been recorded across the region. By mapping record density, species richness and the distribution of selected taxa, the analysis aims to provide a practical, data-informed picture of where wildflowers are most likely to be found.

The report begins with a broad overview of all recorded angiosperms (flowering plants) in the study area, before narrowing its focus to wildflowers. This allows the analysis to show both the overall structure of the dataset and the more specific patterns that are most relevant to people wanting to discover wildflowers on the Lizard.

---

## 🎯 Aims and Objectives

This analysis follows a two-stage approach:

1. To explore patterns in **all recorded angiosperms (flowering plants)**  
2. To then focus more specifically on **wildflowers (herbaceous flowering plants)**  

The key objectives are:

- To map **record density** across the study area  
- To map **species richness** at a 1 km grid scale  
- To summarise **record counts across major taxonomic groups**  
- To identify patterns in recording effort and biodiversity  
- To explore the distribution of selected wildflower species  

---

## 📊 Data Source

All data used in this report are sourced from the [iNaturalist](https://www.inaturalist.org "iNaturalist") API.

The dataset includes:
- Observations of angiosperms (flowering plants)  
- Records within a defined bounding box covering the Lizard Peninsula
- Records meeting “research grade” quality standards  

These filters ensure that the dataset represents a reliable and spatially consistent sample of plant observations.

---

```{r data_load}
source("Incremental update helpers.R")
source("lizard helper functions.R")

# Load your existing saved dataset
load("obs_sf.RData")

# Update it
obs_sf <- update_inat_dataset(
  existing_obs_sf = obs_sf,
  taxon_id = 47125,
  swlat = 49.95,
  swlng = -5.32,
  nelat = 50.10,
  nelng = -5.00,
  max_pages = 100,
  verbose = TRUE
)

# Save the refreshed version
save(obs_sf, file = "obs_sf.RData")

#load GIS shapefile
grid_lizard <- readRDS("GIS Data/grid_lizard.rds")

```

```{r update-taxonomy, echo=FALSE, message=FALSE}

## Update taxonomy for any new taxa
library(dplyr)

# Load existing taxa info if it exists
if (file.exists("all_taxa_inat_info.RData")) {
  load("all_taxa_inat_info.RData")
} else {
  all_taxa_inat_info <- data.frame()
}

# Extract all ancestry IDs from updated dataset
all_taxa_current <- unique(unlist(strsplit(na.omit(obs_sf$taxon.ancestry), "/")))

# Convert to numeric
all_taxa_current <- as.numeric(all_taxa_current)

# Existing taxa IDs
existing_ids <- if (nrow(all_taxa_inat_info) > 0) all_taxa_inat_info$id else numeric(0)

# Find missing taxa
missing_taxa <- setdiff(all_taxa_current, existing_ids)

cat(length(missing_taxa), "new taxa IDs to fetch\n")

# Fetch only missing taxa
if (length(missing_taxa) > 0) {
  
  new_taxa_info <- get_taxa_info(missing_taxa)
  
  # Combine with existing
  all_taxa_inat_info <- bind_rows(all_taxa_inat_info, new_taxa_info) %>%
    distinct(id, .keep_all = TRUE)
  
  # Save updated lookup
  save(all_taxa_inat_info, file = "all_taxa_inat_info.RData")
  
  
}
#high taxo info to inat records
  obs_sf <- add_higher_taxo(obs_sf, all_taxa_inat_info)


```


### Analytical Approach

The analysis is structured in two stages:

#### Stage 1: Angiosperm overview  
All flowering plant records are analysed to understand:
- overall recording patterns  
- spatial distribution of effort  
- broad taxonomic representation  

#### Stage 2: Wildflower focus  
A subset of the data is then used to focus on herbaceous flowering plants (“wildflowers”), allowing more ecologically meaningful interpretation.

---

## 🌿 Angiosperm Overview

This section provides a broad overview of all flowering plant records in the study area.

### Dataset Summary

```{r summary-stats, echo=FALSE, message=FALSE}
library(dplyr)
library(lubridate)
library(knitr)

# Ensure date column is in Date format
obs_sf <- obs_sf %>%
  mutate(
    observed_on = as.Date(observed_on)
  )

# Calculate summary statistics
summary_stats <- list(
  total_records = nrow(obs_sf),
  total_species = n_distinct(obs_sf$taxon.id),
  total_observers = n_distinct(obs_sf$user.login),
  first_record = min(obs_sf$observed_on, na.rm = TRUE),
  last_record = max(obs_sf$observed_on, na.rm = TRUE)
)

# Create a neat table
summary_table <- data.frame(
  Metric = c(
    "Total observations",
    "Total species recorded",
    "Number of observers",
    "Earliest record",
    "Most recent record"
  ),
  Value = c(
    format(summary_stats$total_records, big.mark = ","),
    summary_stats$total_species,
    summary_stats$total_observers,
    as.character(summary_stats$first_record),
    as.character(summary_stats$last_record)
  )
)

kable(summary_table, align = c("l", "r"), col.names = rep("", ncol(summary_table)))
```

#### Record density

```{r leaflet-recording-density, echo=FALSE, message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(leaflet)
library(htmltools)

#--------------------------------------------------
# 1. Ensure both layers use same CRS
#--------------------------------------------------
obs_sf_ll <- st_transform(obs_sf, 4326)
grid_ll   <- st_transform(grid_lizard, 4326)

#--------------------------------------------------
# 2. Join records to grid cells
#--------------------------------------------------
obs_joined <- st_join(obs_sf_ll, grid_ll, join = st_within)

#--------------------------------------------------
# 3. Summarise recording density per grid cell
#--------------------------------------------------
grid_density <- obs_joined %>%
  st_drop_geometry() %>%
  count(tile_name, name = "records_n")

#--------------------------------------------------
# 4. Join counts back to grid
#--------------------------------------------------
grid_map <- grid_ll %>%
  left_join(grid_density, by = "tile_name") %>%
  mutate(
    records_n = ifelse(is.na(records_n), 0, records_n)
  )

#--------------------------------------------------
# 5. Colour palette
#--------------------------------------------------
pal <- colorNumeric(
  palette = "YlOrRd",
  domain = grid_map$records_n[grid_map$records_n > 0],
  na.color = "transparent"
)

#--------------------------------------------------
# 6. Leaflet map
#--------------------------------------------------
leaflet(grid_map) %>%
  
  addProviderTiles(
    providers$CartoDB.Positron
  ) %>%
  
  addPolygons(
    fillColor = ~ifelse(records_n == 0, "transparent", pal(records_n)),
    fillOpacity = ~ifelse(records_n == 0, 0, 0.35),
    color = ~ifelse(records_n == 0, "transparent", "#666666"),
    weight = ~ifelse(records_n == 0, 0, 0.8),
    smoothFactor = 0.2,
    popup = ~ifelse(
      records_n == 0,
      NA,
      paste0(
        "<b>Grid cell:</b> ", tile_name, "<br>",
        "<b>Records:</b> ", format(records_n, big.mark = ",")
      )
    ),
    label = ~ifelse(
      records_n == 0,
      NA,
      paste0(tile_name, ": ", records_n, " records")
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      color = "black",
      fillOpacity = 0.55,
      bringToFront = TRUE
    )
  ) %>%
  
  addLegend(
    pal = pal,
    values = ~records_n[records_n > 0],
    opacity = 0.7,
    title = "Recording density",
    position = "bottomright"
  )
```

---

### Taxonomic Composition

This section summarises record counts across major plant groups (e.g. families or higher taxa).

The aim is to identify:
- which groups are most frequently recorded  
- potential biases in recording effort  
- under-represented plant groups  

```{r sexy-order-barchart-final, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=10}
library(dplyr)
library(ggplot2)
library(scales)
library(forcats)
library(stringr)

#--------------------------------------------------
# 1. Summarise records by order
#--------------------------------------------------
order_summary <- obs_sf %>%
  filter(!is.na(order_common)) %>%
  count(order_common, sort = TRUE, name = "records") %>%
  mutate(order_common = tools::toTitleCase(order_common))

#--------------------------------------------------
# 2. Top 10 + Other
#--------------------------------------------------
top_n <- 10

order_plot <- bind_rows(
  order_summary %>% slice_head(n = top_n),
  tibble(
    order_common = "Other Orders",
    records = sum(order_summary$records[-seq_len(min(top_n, nrow(order_summary)))])
  )
) %>%
  filter(records > 0)

#--------------------------------------------------
# 3. Short labels
#--------------------------------------------------
shorten_order <- function(x) {
  x %>%
    str_replace(", And Allies$", "") %>%
    str_replace("Sunflowers, Bellflowers, Fanflowers", "Sunflowers, Bellflowers") %>%
    str_replace("Mints, Plantains, Olives", "Mints, Plantains") %>%
    str_replace("Agaves, Orchids, Irises", "Agaves, Orchids") %>%
    str_replace("Heathers, Balsams, Primroses", "Heathers, Primroses") %>%
    str_replace("Roses, Elms, Figs", "Roses, Elms") %>%
    str_replace("Honeysuckles, Moschatels", "Honeysuckles") %>%
    str_replace("Saxifrages, Stonecrops", "Saxifrages") %>%
    str_wrap(width = 24)
}

order_plot <- order_plot %>%
  mutate(
    label = shorten_order(order_common),
    label = fct_reorder(label, records)
  )

#--------------------------------------------------
# 4. Plot
#--------------------------------------------------
ggplot(order_plot, aes(x = label, y = records, fill = records)) +
  
  geom_col(
    width = 0.62,   # slimmer bars = more air
    alpha = 0.92
  ) +
  
  geom_text(
    aes(label = comma(records)),
    hjust = -0.15,
    size = 4.2,
    fontface = "bold"
  ) +
  
  coord_flip(clip = "off") +
  
  scale_fill_viridis_c(
    option = "magma",
    direction = -1,
    guide = "none"
  ) +
  
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.16))
  ) +
  
  labs(
    title = "Taxonomic Composition",
    subtitle = "Distribution of iNaturalist observations across major plant orders",
    x = NULL,
    y = "Number of records",
    caption = "Top 10 orders shown individually; remaining orders grouped as Other"
  ) +
  
  theme_minimal(base_size = 13) +
  
  theme(
    plot.title = element_text(
      face = "bold",
      size = 28,
      hjust = 0,
      margin = margin(b = 5)
    ),
    
    plot.subtitle = element_text(
      size = 14,
      colour = "grey35",
      hjust = 0,
      margin = margin(b = 18)
    ),
    
    axis.text.y = element_text(
      size = 11,
      face = "bold",
      lineheight = 0.95
    ),
    
    axis.text.x = element_text(
      size = 10.5,
      colour = "grey35"
    ),
    
    axis.title.x = element_text(
      size = 13,
      face = "bold",
      margin = margin(t = 12)
    ),
    
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(colour = "grey88"),
    
    plot.caption = element_text(
      size = 10,
      colour = "grey45",
      hjust = 0
    ),
    
    plot.margin = margin(10, 40, 10, 10)
  )
```

### Interpretation of Taxonomic Composition

The taxonomic breakdown suggests that the Lizard Peninsula supports a diverse and characteristically coastal flora. The most frequently recorded plant orders include Caryophyllales (“Pinks, Cactuses, and Allies”), Asterales (“Sunflowers, Bellflowers, and Allies”), and Lamiales (“Mints, Plantains, and Allies”). These groups contain many species associated with maritime grassland, heathland, flower-rich verges, rocky ground, and open semi-natural habitats.

The prominence of Caryophyllales is particularly notable, as this order includes species such as thrift, sea campion, common sorrel, knotgrass, and stonecrops. Many of these plants are well adapted to thin soils, exposed coastal conditions, and rocky habitats, which may reflect the distinctive botanical character of the Lizard’s cliffs, heaths, and serpentine landscapes.

High totals for Asterales are also expected. This group includes familiar flowers such as daisies, dandelions, ragworts, knapweeds, thistles, chamomile, and bellflowers. Many are widespread, colourful, and flower for long periods, making them especially likely to be recorded by observers.

Lamiales includes species such as ribwort plantain, speedwells, foxglove, selfheal, eyebrights, dead-nettles, and mints. These plants are commonly associated with grassland, heaths, hedgebanks, and field margins.

Overall, the results indicate that iNaturalist records are capturing genuine ecological patterns rather than being dominated by only a small number of widespread garden or urban species. However, some recorder bias is still likely. Colourful and easily recognised flowers are often over-recorded, while grasses, sedges, rushes, and more difficult plant groups are likely to be under-represented.

---

## 🗺️ Spatial Patterns

### Record Density

Record density is mapped at a 1 km grid scale, showing the number of observations per grid cell.

These patterns are influenced by:
- accessibility  
- observer behaviour  
- landscape features  

---

### Species Richness

Species richness (number of species per grid cell) is mapped to identify areas of high recorded diversity.

These maps highlight:
- potential biodiversity hotspots  
- areas of concentrated recording effort  

---

### Effort and Bias

Differences between record density and species richness are explored to understand how recording effort influences observed patterns.

This helps distinguish between:
- genuinely diverse areas  
- heavily recorded locations  

---

## 🌼 Wildflower Focus

Following the broad angiosperm analysis, this section focuses on wildflowers.

### Defining Wildflowers

“Wildflower” is not a formal taxonomic category. For the purposes of this report, the term refers to herbaceous flowering plants, approximated through taxonomic filtering.

This definition is used to:
- focus on ecologically meaningful species  
- improve interpretability of spatial patterns  

---

### Wildflower Diversity

This section summarises:
- number of wildflower species recorded  
- most frequently observed species  
- distribution across the study area  

---

### Distribution of Selected Species

Selected wildflower species are mapped individually to illustrate:

- habitat associations  
- spatial clustering  
- recording patterns  

These examples help connect the dataset to real-world ecological features of the Lizard Peninsula.

---

## ⚠️ Limitations

Several limitations should be considered:

- iNaturalist data are presence-only and influenced by observer behaviour  
- recording effort is uneven across the study area  
- some plant groups (e.g. grasses and sedges) are under-recorded  
- “wildflower” classification is approximate  

These factors mean that results reflect both biodiversity and recording patterns.

---

## 🔍 Interpretation

The analysis highlights the interaction between:
- ecological diversity  
- landscape accessibility  
- citizen science recording behaviour  

By comparing record density, species richness, and taxonomic composition, it is possible to identify both well-recorded areas and potential gaps.

---

## 🚀 Opportunities for Future Work

This analysis provides a foundation for further work, including:

- refining wildflower classification  
- incorporating plant trait data  
- analysing seasonal trends  
- comparing with other regions  
- supporting targeted survey and engagement activities  

---

## 📣 Conclusion

This study demonstrates how citizen science data can be used to explore plant diversity at a local scale.

By starting with a broad angiosperm dataset and then focusing on wildflowers, it is possible to produce both robust analysis and meaningful ecological insights.

These findings can support:
- biodiversity understanding  
- improved recording strategies  
- public engagement with wildflowers  